Splitting it up: the \pkg{spduration} split-population duration regression package

### 2015-10-28 manuscript.Rmd
writinating the manuscripter

---
author:
  - name: Daniel W. Hill
    affiliation: University of Georgia
    address: >
      First line
      Second line
    email: name@company.com
    url: http://rstudio.com
  - name: Nils W. Metternich
    affiliation: University College London
  - name: Andreas Beger
    affiliation: who knows
  - name: Michael D. Ward
    affiliation: Duke University
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model. The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{R}"]
  plain:     [keywords, not capitalized, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

The `spduration` package implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, the Weibull and log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.} The primary function, \texttt{spdur}, produces a regression model object of class \texttt{spdur} which can then be used with further methods.

To illustrate the usefulness of the package and the broad applicability of the methods outlined above, we employ an example from political science. We replicate and extend the analysis presented in @belkin2003toward, which examines factors that lead to military coups. Belkin and Schofer's distinction between "structural" vs. "triggering"" causes of coups strongly suggests that a split-population approach to this question will be useful. They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors. Using language which fits nicely with the class of models described above, they say "Triggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup. Hence, triggers should not be equated with coup risk. Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [p. 598 @belkin2003toward]. Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups. They employ a logistic regression model which does not adequately capture the process described in the quote above, since it implicitly assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations). Their structural coup risk indicator is developed precisely to distinguish between cases where coups are possible and cases where they are not, so the split-population model allows one to examine whether the indicator effectively separates at-risk cases from cases where coups are impossible.

We begin by loading the package and formatting the data appropriately. 
```{r data}
library("spduration")
#data(bscoup)
library("foreign")
bscoup <- read.dta("data/BelkinSchoferTable4.dta")
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                            freq="year")
```

\normalsize
The \texttt{add}\_\texttt{duration} function takes as input a data frame with a binary response variable (in this example, "coup"") that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named "duration") for each spell as well as a binary variable ("atrisk") that is constant across the spell and indicates whether the spell ended in failure. These are then used as response variables for the formula in the `spdur` function. Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data. We begin by fitting first a Weibull and then a log-logistic split-population duration model using the coups data, including the measure of coup risk in the logit (risk) equation.

@@

The objects produced by \texttt{spdur} can also be used with functions in the \texttt{xtable} package to produce \LaTeX\ code for tables of results. The \texttt{xtable} function used on an \texttt{spdur} object will produce code such as that used to create Table \ref{loglog_table}.

@@

The package also includes a function, \texttt{plot.hazard}, that can be used with an \texttt{spdur} object to plot the estimated hazard rate. \texttt{plot.hazard} plots the conditional hazard, which is the probability of survival conditional on the covariates in the risk and duration equations, and conditional on survival up to time $t$. The function calculates the estimated hazard rate as well as 90\% confidence intervals, which are produced by simulating values from the estimated sampling distributions of the model parameters. By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \texttt{xvals} and \texttt{zvals}, which correspond to the covariates in the duration and risk equations, respectively. The command below creates Figure \ref{plot-hazard2}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
\end{center}
\end{figure}

@@

Importantly, the package also includes functions to evaluate model predictions. The generic function \texttt{predict} can be used on an object of class \texttt{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period. In addition, the generic \texttt{plot} function can be used on \texttt{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions. The code below produces Figure \ref{insamp}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/sepplots.pdf}
\end{center}
\end{figure}

The option \texttt{endSpellOnly} is set to \texttt{FALSE} so that every observation, not only those at the end of a spell, is used in the plot. By default the \texttt{plot} function will calculate the conditional hazard for each observation. The separation plot sorts observations from left to right according to the predicted probability assigned by the model (higher values to the right), and shows each event/failure as red line, with non-events shown in beige. This makes it easy to see whether the model is assigning high probabilities of failure to actual cases of failure, and low probabilities to non-failures.

@@

\normalsize
The \texttt{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set.

@@

\normalsize
The \texttt{train},\texttt{test}, and \texttt{pred} arguments are used to specify data frames for the training set, test set, and forecasting set. We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set. Here we use \texttt{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population. The test set predictions are stored in the model object as \texttt{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \texttt{separationplot} package.

@@

Sometimes information about previous failures in the data is available beyond the time period over which covariates are observed, making it possible to ameliorate or eliminate left-censoring by using the information of previous failures when constructing the necessary duration variables with \texttt{add\_duration()}.

Right-censoring, where spells end before outcomes are observed, also pose a unique problem in the split-population framework. Although right-censored spells themselves are accommodated in the modeling function, they impact the coding of at risk vs. immune spells. The \texttt{add\_duration()} function retroactively codes all observations in a spell as at risk if the spell itself ended in failure. Right-censored spells are coded as immune over their entire duration. This can lead to some misclassification of observations as immune even though they experience failure at some point in the unobserved future.

@@

Data and outcomes in international relations, and to a lesser extent in other fields as well, are typically characterized by temporal dependence and, where a binary, ordinal, or count outcome is modelled, rare events or severely imbalanced classes with excessive numbers of zeros. Figure \ref{rates} shows a few examples for published research that models binary outcomes.

This is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling]. Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics.

In addition to these technical features, split-population duration models are also appealing in the context of theories or arguments that explicitly distinguish long-term structural causes from short-term triggering events, as @belkin2003toward and others do, as this conceptual distinction can be accomodated through the seperate risk and duration equations.

>>>>

---
author:
  - name: Daniel W. Hill
    affiliation: University of Georgia
    address: >
      First line
      Second line
    email: dwhill@uga.edu
    url: http://spia.uga.edu/directory/faculty_staff/hill-daniel
  - name: Nils W. Metternich
    affiliation: University College London
  - name: Andreas Beger
    affiliation: Duke University
  - name: Michael D. Ward
    affiliation: Duke University
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model. The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{R}"]
  plain:     [keywords, not capitalized, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

The \pkg{spduration} package for \proglang{R} implements a split-population duration regression for data with time-varying covariates and with two options for the hazard rate shape, Weibull and log-logistic. The Weibull density allows for hazard rates that are increasing, constant, or decreasing over survival time, while the log-logistic density also can fit rates that have a peak at a particular survival time.

As an example to illustrate the package functionality, we will replicate and extend the model of coup onsets in @belkin2003toward. Belkin and Schofer's paper lends itself to re-analysis with a split-populuation duration model because they explicitly distinguish long-term structural risk factors for coups from more short-term triggering causes that can explain the timing of a coup in a regime at risk. They argue that many countries never experience coups because coups are effectvely impossible due to structural factors, while others that never experience coups are nevertheless at risk due to a different configuration of those same factors. Using language which fits nicely with the class of models described above, they say "Triggers are not the source of the original risk, and in the absence of structural causes, the presence of triggering factors alone cannot lead to a coup. Hence, triggers should not be equated with coup risk. Rather, they are factors that may determine the exact timing of a coup in regimes that suffer from high coup risk" [p. 598 @belkin2003toward]

Belkin and Schofer develop a measure of ``structural coup risk'' which incorporates factors such as the strength of democratic institutions and civil society, and a recent history of successful coups. They employ a logistic regression model which does not adequately capture the process described in the quote above, since it implicitly assumes that all observations are at risk for a coup (i.e., the probability of a coup is non-zero for all observations). Their structural coup risk indicator is developed precisely to distinguish between cases where coups are possible and cases where they are not, so the split-population model allows one to examine whether the indicator effectively separates at-risk cases from cases where coups are impossible.

We begin by loading the package and the Belkin and Schofer replication data, and formatting the data to add several variables needed by the split-population duration model.

```{r data}
library("spduration")
#data(bscoup)
library("foreign")
bscoup <- read.dta("data/BelkinSchoferTable4.dta")
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                            freq="year")
```

\normalsize
The \code{add\_duration} function takes as input a data frame with a binary response variable (in this example, "coup"") that measures the occurrence of the event, or failure, recorded over discrete time periods, and produces a data frame with a duration variable (named "duration") for each spell as well as a binary variable ("atrisk") that is constant across the spell and indicates whether the spell ended in failure. These are then used as response variables for the formula in the `spdur` function. The \code{spdur} function is the primary function in the package and produces a regression model object of class \code{spdur} which can then be used with further methods.

Though the data used in this example are recorded annually, the function supports annual, monthly, or daily data. We begin by fitting first a Weibull and then a log-logistic split-population duration model using the coups data, including the measure of coup risk in the logit (risk) equation.

@@

The objects produced by \code{spdur} can also be used with functions in the \code{xtable} package to produce \LaTeX\ code for tables of results. The \code{xtable} function used on an \code{spdur} object will produce code such as that used to create Table \ref{loglog_table}.

@@

The package also includes a function, \code{plot_hazard}, that can be used with an \code{spdur} object to plot the estimated hazard rate. \code{plot_hazard} plots the conditional hazard, which is the probability of survival conditional on the covariates in the risk and duration equations, and conditional on survival up to time $t$. The function calculates the estimated hazard rate as well as 90\% confidence intervals, which are produced by simulating values from the estimated sampling distributions of the model parameters. By default the function uses the mean values of the covariates during the simulations, but users can choose specific covariate values by entering them as vectors in the arguments \code{xvals} and \code{zvals}, which correspond to the covariates in the duration and risk equations, respectively. The command below creates the graph A in Figure \ref{hazard-ex}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/hazard-ex.pdf}
\caption{Plots of the hazard rate for the log-logistic model of coups. The left graph uses the default mean values for covariates, while graph B uses user-specified variable values for a high-risk military regime.}
\label{hazard-ex}
\end{center}
\end{figure}

@@

Importantly, the package also includes functions to evaluate model predictions. The generic function \code{predict} can be used on an object of class \code{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period. In addition, the generic \code{plot} function can be used on \code{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions. The code below produces Figure \ref{insamp}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{graphics/sepplots.pdf}
\caption{In-sample separation plots of Weibull and log-logistic model conditional hazard predictions}
\label{insamp}
\end{center}
\end{figure}

The option \code{endSpellOnly} is set to \code{FALSE} so that every observation, not only those at the end of a spell, is used in the plot. By default the \code{plot} function will calculate the conditional hazard for each observation. The separation plot sorts observations from left to right according to the predicted probability assigned by the model (higher values to the right), and shows each event/failure as red line, with non-events shown in beige. This makes it easy to see whether the model is assigning high probabilities of failure to actual cases of failure, and low probabilities to non-failures.

@@

\normalsize
The \code{spdurCrisp} function can then be used to fit a model on the training set and generate predictions for the test set.

@@

\normalsize
The \code{train},\code{test}, and \code{pred} arguments are used to specify data frames for the training set, test set, and forecasting set. We do not wish to generate forecasts here, so we simply specify the same data frame for the test and forecasting set. Here we use \code{spdurCrisp} to calculate the conditional risk, which is the default calculation for this function and is the probability that an observation is in the at-risk population. The test set predictions are stored in the model object as \code{test.p}, and we extract them to produce the plot shown below in Figure \ref{outsamp}, which was created using the \code{separationplot} package.

@@

Sometimes information about previous failures in the data is available beyond the time period over which covariates are observed, making it possible to ameliorate or eliminate left-censoring by using the information of previous failures when constructing the necessary duration variables with \code{add\_duration()}.

Right-censoring, where spells end before outcomes are observed, also pose a unique problem in the split-population framework. Although right-censored spells themselves are accommodated in the modeling function, they impact the coding of at risk vs. immune spells. The \code{add\_duration()} function retroactively codes all observations in a spell as at risk if the spell itself ended in failure. Right-censored spells are coded as immune over their entire duration. This can lead to some misclassification of observations as immune even though they experience failure at some point in the unobserved future.

@@

Many outcomes in the area that we work in, political science and political violence more specifically, are rare events. Coups, the onset of war and civil war, mass killings, and so on are exceedingly rare events when one considers that most countries most of the time are stable. Figure \ref{rates} shows a few examples for published research that models binary outcomes. @fearon2003ethnicity is a very widely cited study of civil war onset that uses yearly observations of all major countries, and @beger2015ensemble is an example of the positive rates when moving to monthly data for a similar set of countries.

In the language of machine learning, we are dealing with highly imbalanced classes, although our data are also highly structured with clear dependence over space and time in countries and societies that may experience political violence. This is a well-recognized problem and has led to the development or use of several specialized mixture models like zero-inflated Poisson and negative binomial regression for count data, and a zero-inflated ordered probit for ordinal outcomes [@bagozzi2015modeling]. Split-population duration regression provides another principled solution to the challenges posed by data in this domain, but, unlike other solutions to the sparse outcome problem, also addresses underlying temporal dynamics.

Split-population duration models are not only appealing in this technical sense, but they also match the logic or intuituion many analysts of politics use when they distinguish long-term risk factors from more fleeting triggering causes. The example we have used, @belkin2003toward, is a particularly clear illustration of how well the language of theorists maps onto the model intuition.

### 2015-10-28 manuscript.tex
writinating the manuscripter

\author{
Daniel W. Hill\\University of Georgia \And Nils W. Metternich\\University College London \And Andreas Beger\\who knows \And Michael D. Ward\\Duke University
}
\title{Splitting it up: the \pkg{spduration} split-population duration
regression package}
\Keywords{keywords, not capitalized, \proglang{R}}

@@

\Address{
    Daniel W. Hill\\
  University of Georgia\\
  First line Second line\\
  E-mail: \href{mailto:name@company.com}{\nolinkurl{name@company.com}}\\
  URL: \url{http://rstudio.com}\\~\\
        }

@@

The \texttt{spduration} package implements a split-population duration
regression for data with time-varying covariates and with two options
for the hazard rate shape, the Weibull and
log-logistic.\footnote{The Weibull allows for monotonically increasing or decreasing hazard rates, while the log-logistic allows for rates that first increase and then decrease.}
The primary function, \texttt{spdur}, produces a regression model object
of class \texttt{spdur} which can then be used with further methods.

To illustrate the usefulness of the package and the broad applicability
of the methods outlined above, we employ an example from political
science. We replicate and extend the analysis presented in Belkin and
Schofer (2003), which examines factors that lead to military coups.
Belkin and Schofer's distinction between ``structural'' vs.
``triggering''" causes of coups strongly suggests that a
split-population approach to this question will be useful. They argue
that many countries never experience coups because coups are effectvely
impossible due to structural factors, while others that never experience
coups are nevertheless at risk due to a different configuration of those
same factors. Using language which fits nicely with the class of models
described above, they say ``Triggers are not the source of the original
risk, and in the absence of structural causes, the presence of
triggering factors alone cannot lead to a coup. Hence, triggers should
not be equated with coup risk. Rather, they are factors that may
determine the exact timing of a coup in regimes that suffer from high
coup risk'' (p.~598 Belkin and Schofer 2003). Belkin and Schofer develop
a measure of ``structural coup risk'' which incorporates factors such as
the strength of democratic institutions and civil society, and a recent
history of successful coups. They employ a logistic regression model
which does not adequately capture the process described in the quote
above, since it implicitly assumes that all observations are at risk for
a coup (i.e., the probability of a coup is non-zero for all
observations). Their structural coup risk indicator is developed
precisely to distinguish between cases where coups are possible and
cases where they are not, so the split-population model allows one to
examine whether the indicator effectively separates at-risk cases from
cases where coups are impossible.

We begin by loading the package and formatting the data appropriately.

@@

\normalsize
The \texttt{add}\_\texttt{duration} function takes as input a data frame
with a binary response variable (in this example, ``coup''``) that
measures the occurrence of the event, or failure, recorded over discrete
time periods, and produces a data frame with a duration variable
(named''duration``) for each spell as well as a binary variable
(''atrisk``) that is constant across the spell and indicates whether the
spell ended in failure. These are then used as response variables for
the formula in the \texttt{spdur} function. Though the data used in this
example are recorded annually, the function supports annual, monthly, or
daily data. We begin by fitting first a Weibull and then a log-logistic
split-population duration model using the coups data, including the
measure of coup risk in the logit (risk) equation.

@@

The objects produced by \texttt{spdur} can also be used with functions
in the \texttt{xtable} package to produce \LaTeX~code for tables of
results. The \texttt{xtable} function used on an \texttt{spdur} object
will produce code such as that used to create Table \ref{loglog_table}.

@@

The package also includes a function, \texttt{plot.hazard}, that can be
used with an \texttt{spdur} object to plot the estimated hazard rate.
\texttt{plot.hazard} plots the conditional hazard, which is the
probability of survival conditional on the covariates in the risk and
duration equations, and conditional on survival up to time \(t\). The
function calculates the estimated hazard rate as well as 90\% confidence
intervals, which are produced by simulating values from the estimated
sampling distributions of the model parameters. By default the function
uses the mean values of the covariates during the simulations, but users
can choose specific covariate values by entering them as vectors in the
arguments \texttt{xvals} and \texttt{zvals}, which correspond to the
covariates in the duration and risk equations, respectively. The command
below creates Figure \ref{plot-hazard2}.

@@

Importantly, the package also includes functions to evaluate model
predictions. The generic function \texttt{predict} can be used on an
object of class \texttt{spdur} to generate several kinds of predictions,
including the probability that an observation is ``at-risk'' and the
probability of failure for a given time period. In addition, the generic
\texttt{plot} function can be used on \texttt{spdur} objects to produce
a separation plot (Greenhill, Ward, and Sacks 2011), which is a
graphical display for evaluating model predictions. The code below
produces Figure \ref{insamp}.

@@

Sometimes information about previous failures in the data is available
beyond the time period over which covariates are observed, making it
possible to ameliorate or eliminate left-censoring by using the
information of previous failures when constructing the necessary
duration variables with \texttt{add\_duration()}.

Right-censoring, where spells end before outcomes are observed, also
pose a unique problem in the split-population framework. Although
right-censored spells themselves are accommodated in the modeling
function, they impact the coding of at risk vs.~immune spells. The
\texttt{add\_duration()} function retroactively codes all observations
in a spell as at risk if the spell itself ended in failure.
Right-censored spells are coded as immune over their entire duration.
This can lead to some misclassification of observations as immune even
though they experience failure at some point in the unobserved future.

@@

Data and outcomes in international relations, and to a lesser extent in
other fields as well, are typically characterized by temporal dependence
and, where a binary, ordinal, or count outcome is modelled, rare events
or severely imbalanced classes with excessive numbers of zeros. Figure
\ref{rates} shows a few examples for published research that models
binary outcomes.

This is a well-recognized problem and has led to the development or use
of several specialized mixture models like zero-inflated Poisson and
negative binomial regression for count data, and a zero-inflated ordered
probit for ordinal outcomes (Bagozzi et al. 2015). Split-population
duration regression provides another principled solution to the
challenges posed by data in this domain, but, unlike other solutions to
the sparse outcome problem, also addresses underlying temporal dynamics.

In addition to these technical features, split-population duration
models are also appealing in the context of theories or arguments that
explicitly distinguish long-term structural causes from short-term
triggering events, as Belkin and Schofer (2003) and others do, as this
conceptual distinction can be accomodated through the seperate risk and
duration equations.

>>>>

\author{
Daniel W. Hill\\University of Georgia \And Nils W. Metternich\\University College London \And Andreas Beger\\Duke University \And Michael D. Ward\\Duke University
}
\title{Splitting it up: the \pkg{spduration} split-population duration
regression package}
\Keywords{keywords, not capitalized, \proglang{R}}

@@

\Address{
    Daniel W. Hill\\
  University of Georgia\\
  First line Second line\\
  E-mail: \href{mailto:dwhill@uga.edu}{\nolinkurl{dwhill@uga.edu}}\\
  URL: \url{http://spia.uga.edu/directory/faculty_staff/hill-daniel}\\~\\
        }

@@

The \pkg{spduration} package for \proglang{R} implements a
split-population duration regression for data with time-varying
covariates and with two options for the hazard rate shape, Weibull and
log-logistic. The Weibull density allows for hazard rates that are
increasing, constant, or decreasing over survival time, while the
log-logistic density also can fit rates that have a peak at a particular
survival time.

As an example to illustrate the package functionality, we will replicate
and extend the model of coup onsets in Belkin and Schofer (2003). Belkin
and Schofer's paper lends itself to re-analysis with a split-populuation
duration model because they explicitly distinguish long-term structural
risk factors for coups from more short-term triggering causes that can
explain the timing of a coup in a regime at risk. They argue that many
countries never experience coups because coups are effectvely impossible
due to structural factors, while others that never experience coups are
nevertheless at risk due to a different configuration of those same
factors. Using language which fits nicely with the class of models
described above, they say ``Triggers are not the source of the original
risk, and in the absence of structural causes, the presence of
triggering factors alone cannot lead to a coup. Hence, triggers should
not be equated with coup risk. Rather, they are factors that may
determine the exact timing of a coup in regimes that suffer from high
coup risk'' (p.~598 Belkin and Schofer 2003)

Belkin and Schofer develop a measure of ``structural coup risk'' which
incorporates factors such as the strength of democratic institutions and
civil society, and a recent history of successful coups. They employ a
logistic regression model which does not adequately capture the process
described in the quote above, since it implicitly assumes that all
observations are at risk for a coup (i.e., the probability of a coup is
non-zero for all observations). Their structural coup risk indicator is
developed precisely to distinguish between cases where coups are
possible and cases where they are not, so the split-population model
allows one to examine whether the indicator effectively separates
at-risk cases from cases where coups are impossible.

We begin by loading the package and the Belkin and Schofer replication
data, and formatting the data to add several variables needed by the
split-population duration model.

@@

\normalsize
The \code{add\_duration} function takes as input a data frame with a
binary response variable (in this example, ``coup''``) that measures the
occurrence of the event, or failure, recorded over discrete time
periods, and produces a data frame with a duration variable
(named''duration``) for each spell as well as a binary variable
(''atrisk``) that is constant across the spell and indicates whether the
spell ended in failure. These are then used as response variables for
the formula in the \texttt{spdur} function. The \code{spdur} function is
the primary function in the package and produces a regression model
object of class \code{spdur} which can then be used with further
methods.

Though the data used in this example are recorded annually, the function
supports annual, monthly, or daily data. We begin by fitting first a
Weibull and then a log-logistic split-population duration model using
the coups data, including the measure of coup risk in the logit (risk)
equation.

@@

The objects produced by \code{spdur} can also be used with functions in
the \code{xtable} package to produce \LaTeX~code for tables of results.
The \code{xtable} function used on an \code{spdur} object will produce
code such as that used to create Table \ref{loglog_table}.

@@

The package also includes a function, \code{plot_hazard}, that can be
used with an \code{spdur} object to plot the estimated hazard rate.
\code{plot_hazard} plots the conditional hazard, which is the
probability of survival conditional on the covariates in the risk and
duration equations, and conditional on survival up to time \(t\). The
function calculates the estimated hazard rate as well as 90\% confidence
intervals, which are produced by simulating values from the estimated
sampling distributions of the model parameters. By default the function
uses the mean values of the covariates during the simulations, but users
can choose specific covariate values by entering them as vectors in the
arguments \code{xvals} and \code{zvals}, which correspond to the
covariates in the duration and risk equations, respectively. The command
below creates the graph A in Figure \ref{hazard-ex}.

@@

Importantly, the package also includes functions to evaluate model
predictions. The generic function \code{predict} can be used on an
object of class \code{spdur} to generate several kinds of predictions,
including the probability that an observation is ``at-risk'' and the
probability of failure for a given time period. In addition, the generic
\code{plot} function can be used on \code{spdur} objects to produce a
separation plot (Greenhill, Ward, and Sacks 2011), which is a graphical
display for evaluating model predictions. The code below produces Figure
\ref{insamp}.

@@

Sometimes information about previous failures in the data is available
beyond the time period over which covariates are observed, making it
possible to ameliorate or eliminate left-censoring by using the
information of previous failures when constructing the necessary
duration variables with \code{add\_duration()}.

Right-censoring, where spells end before outcomes are observed, also
pose a unique problem in the split-population framework. Although
right-censored spells themselves are accommodated in the modeling
function, they impact the coding of at risk vs.~immune spells. The
\code{add\_duration()} function retroactively codes all observations in
a spell as at risk if the spell itself ended in failure. Right-censored
spells are coded as immune over their entire duration. This can lead to
some misclassification of observations as immune even though they
experience failure at some point in the unobserved future.

@@

Many outcomes in the area that we work in, political science and
political violence more specifically, are rare events. Coups, the onset
of war and civil war, mass killings, and so on are exceedingly rare
events when one considers that most countries most of the time are
stable. Figure \ref{rates} shows a few examples for published research
that models binary outcomes.

In the language of machine learning, we are dealing with highly
imbalanced classes, although our data are also highly structured with
clear dependence over space and time in countries and societies that may
experience political violence. This is a well-recognized problem and has
led to the development or use of several specialized mixture models like
zero-inflated Poisson and negative binomial regression for count data,
and a zero-inflated ordered probit for ordinal outcomes (Bagozzi et al.
2015). Split-population duration regression provides another principled
solution to the challenges posed by data in this domain, but, unlike
other solutions to the sparse outcome problem, also addresses underlying
temporal dynamics.

Split-population duration models are not only appealing in this
technical sense, but they also match the logic or intuituion many
analysts of politics use when they distinguish long-term risk factors
from more fleeting triggering causes. The example we have used, Belkin
and Schofer (2003), is a particularly clear illustration of how well the
language of theorists maps onto the model intuition.
